---
title: "Mini Snorkel Feather HSC Using Hurdle Model"
author: "Maddee Rubenson"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: false
    math_method:
      engine: webtex
      url: https://latex.codecogs.com/svg.image?
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
    math_method:
      engine: webtex
      url: https://latex.codecogs.com/svg.image?
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(tidymodels)

knitr::opts_chunk$set(echo = TRUE)

theme_set(
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),  
    axis.title.x = element_text(size = 14),  
    axis.title.y = element_text(size = 14)   
  )
)
options(scipen=999)


```

## Objective

Using the mini snorkel data for the Feather River, produce an HSC that
uses cover, substrate, depth and velocity.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# read in mini snorkel data
# TODO: this will get updated to read from EDI when the data package is published
micro_hab <- read_csv(here::here('data-raw', 'microhabitat_observations.csv'))
locations <- read_csv(here::here('data-raw', 'survey_locations.csv'))

mini_snorkel_raw <- micro_hab |> left_join(locations) |> 
  mutate(count = ifelse(is.na(count), 0, count),
        fish_presence = as.factor(ifelse(count < 1, "0", "1"))) |> 
  glimpse()

```

## Data Exploration 

### Outliers 

* There are two outliers, one in the high flow and one in the low flow channel 

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Are the outliers in high flow or low flow channel?
mini_snorkel_raw |> 
  ggplot() + 
  geom_point(aes(x = channel_location, y = count))

mini_snorkel_raw |> 
  filter(count > 0) |> 
  ggplot() + 
  geom_histogram(aes(count))

# remove two outliers 
mini_snorkel_model_ready <- mini_snorkel_raw 
```

```{r echo=FALSE, message=FALSE, warning=FALSE}

# looking at overall count data between the low flow/high flow channels
mini_snorkel_model_ready |> 
  group_by(channel_location) |> 
  tally(count) |> 
  knitr::kable(caption = "Table 1. Total count of fish between high flow and low flow channels")

# This figure shows how there are more fish present in the high flow channel 
# for less time. The fish stick around for much longer in the low flow channel. 
mini_snorkel_model_ready |> 
  group_by(channel_location, month = lubridate::month(date)) |> 
  tally(count) |> 
  ggplot() + 
  geom_col(aes(x = as.factor(month), y = n, fill = channel_location), position = "dodge") + 
  ylab("fish count") +
  xlab("month") +
  ggtitle('Figure 1. Fish presence as a functon of month and high flow/low flow channel')
  
```

# Hurdle Models and Interpretation

A hurdle model was used in Gard 2024 to test for the effects of cover
and habitat type on the total abundance of Chinook salmon at both site
and cell level. Here we use the hurdle model to help understand the
influence of velocity, depth, and cover on fish count and
presence/absence.

**Hurdle Models**

Hurdle models are used when count data has an excess of zeros. These
models can be understood as a mixture of two subset of populations. In
one subset, we have a usual count model that may or may not generate
zero, and the other subset only produce zero count.

A hurdle model models excess zeroes separately from the rest of the
data. The zero counts are modeled as a binary response variable and the
positive counts are modeled using poisson distribution.

*Interpreting a Hurdle Model*

The binary part of the model helps identify factors that influence the
presence/absence of fish. The coefficients of the zero part of the
hurdle model represent the odds ratio of observing at least one fish.

The count part of the model estimate the effects of predictor variables
on the count outcome, excluding all zero counts. Coefficients of counts
represent rate ratios of one or more fish observed.

The Incidence Result Ratio (IRR) in the count part of the model (count
\> 0) represent the multiplicative effect of a one-unit change in a
predictor variable on the expected count of non-zero observations,
assuming all other variables are held constant. For example, if the IRR
for a predictor is 1.2, it means that a one-unit increase in that
predictor is associated with a 20% increase in the expected count of
non-zero observations, assuming all other variables remain constant. For
the binary part of the model - if the coefficient for a predictor in the
binary part of the hurdle model is 0.5, it means that a one-unit
increase in the predictor is associated with a 50% increase in the odds
of having a zero count versus a positive count, assuming all other
variables are held constant.

## Binary Cover Values

**Notes** For this hurdle model, chose to make all substrate and cover
variables presence/absence based on a 20% threshold

**Predictors**

-   small woody inchannel + half meter overhead
-   small woody inchannel + more than half meter overhead
-   large woody inchannel + half meter overhead
-   large woody inchannel + more than half meter overhead
-   cobble substrate
-   boulder substrate
-   small woody inchannel
-   large woody inchannel
-   submerged aquatic veg inchannel
-   undercut bank
-   more than half meter overhead
-   small woody half meter overhead
-   small woody more than half meter overhead
-   velocity
-   depth
-   channel_location (high flow, low flow)

```{r echo=FALSE, message=FALSE, warning=FALSE}
percent_threshold <- 20

# use a hurdle model (Gard 2024); 
# pre-processing - all numeric values must be rounded to whole numbers
hurdle_data <- mini_snorkel_model_ready |>
    select(count,channel_location, depth, velocity, contains("inchannel"), contains("overhead"), "percent_cobble_substrate", "percent_boulder_substrate", "percent_undercut_bank") |>
  # create new cover variables, based off Gard 2023
  mutate(small_woody_inchannel_and_half_meter_overhead = percent_cover_half_meter_overhead + percent_small_woody_cover_inchannel,
         small_woody_inchannel_and_more_than_half_meter_overhead = percent_cover_more_than_half_meter_overhead + percent_small_woody_cover_inchannel,
         large_woody_inchannel_and_half_meter_overhead = percent_cover_half_meter_overhead + percent_large_woody_cover_inchannel,
         large_woody_inchannel_and_more_than_half_meter_overhead = percent_cover_more_than_half_meter_overhead + percent_large_woody_cover_inchannel) |>
  # transform to presence/absence based on a defined threshold
  mutate(cobble_substrate = ifelse(percent_cobble_substrate >= percent_threshold, 1, 0 ),
         boulder_substrate = ifelse(percent_boulder_substrate >= percent_threshold, 1, 0 ),
         small_woody_cover_inchannel = ifelse(percent_small_woody_cover_inchannel >= percent_threshold, 1, 0 ),
         large_woody_cover_inchannel = ifelse(percent_large_woody_cover_inchannel >= percent_threshold, 1, 0 ),
         submerged_aquatic_veg_inchannel = ifelse(percent_submerged_aquatic_veg_inchannel >= percent_threshold, 1, 0),
         undercut_bank = ifelse(percent_undercut_bank >= percent_threshold, 1, 0),
         no_cover_overhead = ifelse(percent_no_cover_overhead >= percent_threshold, 1, 0),
         half_meter_overhead = ifelse(percent_cover_half_meter_overhead >= percent_threshold, 1, 0),
         cover_more_than_half_meter_overhead = ifelse(percent_cover_more_than_half_meter_overhead >= percent_threshold, 1, 0),
         no_cover_inchannel = ifelse(percent_no_cover_inchannel >= percent_threshold, 1, 0),
         small_woody_half_meter_overhead = ifelse(small_woody_inchannel_and_half_meter_overhead >= percent_threshold, 1, 0),
         small_woody_cover_more_than_half_meter_overhead = ifelse(small_woody_inchannel_and_more_than_half_meter_overhead >= percent_threshold, 1, 0),
         large_woody_inchannel_and_half_meter_overhead = ifelse(large_woody_inchannel_and_more_than_half_meter_overhead >= percent_threshold, 1, 0),
         large_woody_cover_more_than_half_meter_overhead = ifelse(large_woody_inchannel_and_more_than_half_meter_overhead >= percent_threshold, 1, 0)) |>
  select(count, depth, velocity, cobble_substrate:large_woody_cover_more_than_half_meter_overhead, channel_location) |>
  # remove no cover variables (no cover overhead, no cover in channel)
  select(-contains("no_cover")) |>
  #mutate_all(~ round(., digits = 0)) |>
  na.omit() |>
  glimpse()


hurdle_model <- pscl::hurdle(count ~ small_woody_cover_inchannel + depth + velocity + large_woody_cover_inchannel + submerged_aquatic_veg_inchannel + cover_more_than_half_meter_overhead + half_meter_overhead + cobble_substrate + boulder_substrate + undercut_bank + channel_location, data = hurdle_data, dist = "negbin") 
# Negative Binomial Distribution: The Negative Binomial distribution is more flexible than the Poisson distribution and is used when the variance of the counts is greater than the mean, which is known as overdispersion. The Negative Binomial model allows for the variance to be larger than the mean, which is common in count data where there is extra variability beyond what is expected from a Poisson distribution.


# zero inflated models are capable of dealing with excess zero counts
# NOTE: Chose to not use a zero inflated model because it operates under the assumption that excess zeros are generated by a separate process from the count values. 
# zeroinf_model <- pscl::zeroinfl(count ~ ., data = hurdle_data) # , zero.dist = "binomial", dist = "negbin"

best_model_hurdle <- MASS::stepAIC(hurdle_model, trace = TRUE)

summary(best_model_hurdle)

```

### Results

```{r echo=FALSE}
library(sjPlot)

# tab_model(best_model_hurdle)

plot_model(best_model_hurdle, title = "Hurdle Model Results Using Binary Presence/Absence for Cover and Substrate")

```

## Using Absolute Values

This hurdle model uses substrate and cover percent values.

**Predictors**

-   depth

-   velocity

-   percent large woody cover in channel

-   percent small woody cover in channel

-   percent cover half meter overhead

-   percent cover more than half meter overhead

-   percent cobble

-   percent boulder

-   percent undercut bank

-   channel_location (high flow, low flow)

**Notes** \*

-   removed combined cover variables (small and large woody in channel
    with cover overhead) because of issues with collinearity within the
    hurdle model

```{r echo=FALSE, message=FALSE, warning=FALSE}

# use a hurdle model (Gard 2024); 
# pre-processing - all numeric values must be rounded to whole numbers

hurdle_data <- mini_snorkel_model_ready |>
    select(count, depth, velocity, channel_location, contains("inchannel"), contains("overhead"), "percent_cobble_substrate", "percent_boulder_substrate", "percent_undercut_bank") |>
  # create new cover variables, based off Gard 2023
  mutate(small_woody_inchannel_and_half_meter_overhead = percent_cover_half_meter_overhead + percent_small_woody_cover_inchannel,
         small_woody_inchannel_and_more_than_half_meter_overhead = percent_cover_more_than_half_meter_overhead + percent_small_woody_cover_inchannel,
         large_woody_inchannel_and_half_meter_overhead = percent_cover_half_meter_overhead + percent_large_woody_cover_inchannel,
         large_woody_inchannel_and_more_than_half_meter_overhead = percent_cover_more_than_half_meter_overhead + percent_large_woody_cover_inchannel) |>
  # remove no cover variables (no cover overhead, no cover in channel)
  select(-contains("no_cover")) |>
  #mutate_all(~ round(., digits = 0)) |>
  na.omit() |>
  distinct() |>
  glimpse()

hurdle_model <- pscl::hurdle(count ~ percent_small_woody_cover_inchannel + depth + velocity + percent_large_woody_cover_inchannel + percent_submerged_aquatic_veg_inchannel + percent_cover_more_than_half_meter_overhead + percent_cover_half_meter_overhead + percent_cobble_substrate + percent_boulder_substrate + percent_undercut_bank + channel_location, data = hurdle_data, dist = "negbin") 
# Negative Binomial Distribution: The Negative Binomial distribution is more flexible than the Poisson distribution and is used when the variance of the counts is greater than the mean, which is known as overdispersion. The Negative Binomial model allows for the variance to be larger than the mean, which is common in count data where there is extra variability beyond what is expected from a Poisson distribution.

best_model_hurdle <- MASS::stepAIC(hurdle_model, trace = TRUE)

summary(best_model_hurdle)

```

### Results

```{r echo=FALSE, message=FALSE, warning=FALSE}
# tab_model(best_model_hurdle)

plot_model(best_model_hurdle, title = "Hurdle Model Results Using Absolute Values")

```

```{r}
knitr::knit_exit()
```

## Next steps

1.  Explore using a fixed effect - such as geomorphic unit (pool, glide,
    riffle) or river mile
