---
title: "Mini Snorkel Feather HSI Workflow"
author: "Maddee Rubenson (FlowWest)"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(tidymodels)

knitr::opts_chunk$set(echo = TRUE)

theme_set(theme_minimal())

```

```{r}
# read in mini snorkel data
# mini_snorkel <- read_csv('../../feather-mini-snorkel/data/microhabitat_with_fish_observations.csv') |> glimpse()

mini_snorkel <- read_csv('https://raw.githubusercontent.com/FlowWest/feather-mini-snorkel/qc-data/data/microhabitat_with_fish_observations.csv?token=GHSAT0AAAAAACCSPYJQGEYKAVJUKVZ5IMEOZQMOPHQ') |> glimpse()

```

## Draft Workflow

[**2004a**](https://netorg629193.sharepoint.com/:b:/s/VA-FeatherRiver/EY9qLwY15ypFn3W42E02takBH745JkefSPAbS4y9VFzuBQ?e=n2QRXK)
Data Analysis

-   Stepwise binary logistic regression analysis was used to assess
    factors influencing the occurrence of steelhead and Chinook salmon.
    Fine scale survey results were analyzed at both the mesohabitat (100
    m2) and microhabitat (1 m2) scale.

-   Mesohabitat analysis was performed by treating the entire 25 m reach
    as a sample

-   Reach habitat variables where steelhead or salmon were present
    (logistic response variable) were compared to reaches where fish
    were absent (logistic reference variable)

-   Microhabitat was done similarly except that individual one square
    meter cells were considered rather than entire reaches

-   Reaches lacking salmon or steelhead were not included in
    microhabitat analysis

[**2005**](https://netorg629193.sharepoint.com/:b:/s/VA-FeatherRiver/ES8H3f7ZUO5Aj7OK8hfW03UB721tCgZd_OEb8P9cbJLiMA?e=L95bf8):
HSC development- chinook salmon rearing

-   HSC were created for fry (\<50 mm) and juvenile Chinook salmon depth
    and mean column velocity data using a three-point running mean to
    smooth frequency distributions of the fry habitat use data and using
    NPTL [what does this stand for?] for the juvenile habitat use data.

-   all substrate suitability given value of 1 after finding that
    substrate was not driving factor for microhabitat selection of fry
    and juveniles

-   cover modified to: with and without. Suitability of without cover
    was calculated as the percentage of fish observed without cover to
    the total sample size

-   suitability of cover present was assigned a value of 1.0 and 0.30 or
    0.22 for cover absent for Chinook salon fry and juveniles [I do not
    get this... ]

-   Focus on suitability of instream cover. When a cover variable did
    not exist, preference for low velocity and shallow depth in a large
    river indicate suitable habitat along stream margins or out in the
    main channel when the river is nearly dry and the preferred
    conditions are prevalent.

-   intermediate scale data separated into four cover types: no cover,
    object only cover overhead only cover, both object and overhead
    cover

[**Gard
2023**](https://netorg629193.sharepoint.com/:b:/s/VA-FeatherRiver/EelfImRfhzxKjLQdrstbIPgBqbRV5S0Wke5GkSh06_CUrQ?e=6PdWKh) -
HSC comparison

-   Presence/absence HSC are developed using a polynomial logistic
    regression that uses both the occupied and unoccupied data; the
    results are then rescaled that the highest value is 1 to calculate
    the HSI

-   For depth and velocity HSC, the criteria were developed directly
    from use observations using a range of curve fitting and smoothing
    techniques

-   Use/availability criteria are developed by dividing use
    observations, generally binned, by availability data from transects

```{r}

mini_snorkel_model_ready <- mini_snorkel |> 
  select(depth:percent_cover_more_than_half_meter_overhead, count, channel_geomorphic_unit, species) |> 
  mutate(fish_presence = ifelse(count < 1, 0, 1)) # fish are always present in this dataset 

```


