---
title: "Mini Snorkel Feather HSC Using Logistic Regression"
author: "Maddee Rubenson"
date: "`r Sys.Date()`"
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(tidymodels)

knitr::opts_chunk$set(echo = TRUE)

theme_set(theme_minimal())

options(scipen=999)


```

```{r}
# read in mini snorkel data
mini_snorkel_model_ready <- read_csv('microhabitat_with_fish_observations.csv') |> 
  mutate(count = ifelse(is.na(count), 0, count),
        fish_presence = as.factor(ifelse(count < 1, "0", "1"))) |> 
  glimpse()

```

## Pre-process data

#### Normalize Substrate by Prevalence

This table provides a weighting for each substrate type based on the
overall presence (\>20%) of each substrate type. Use this to normalize
the substrate columns.

```{r}
substrate_percent <- mini_snorkel_model_ready |> 
  group_by(micro_hab_data_tbl_id) |> 
  select(contains('substrate')) |> 
  distinct() |> 
  pivot_longer(cols = c(percent_fine_substrate:percent_boulder_substrate), names_to = "substrate_type", values_to = "percent") |> 
  mutate(substrate_presence_absence = ifelse(percent < 19, 0, 1)) |>  # 20% threshold
  group_by(substrate_type) |> 
  summarise(total_presence = sum(substrate_presence_absence)) |> 
  ungroup() |> 
  mutate(perc_total = total_presence/sum(total_presence))

knitr::kable(substrate_percent |> mutate(perc_total = perc_total*100), digits = 2)

```

Apply substrate normalization values to substrate columns and remove
unnecessary columns

```{r}
mini_snorkel_grouped <- mini_snorkel_model_ready |> 
  rowwise() |> 
  mutate(fine_substrate = percent_fine_substrate * (1-substrate_percent[substrate_percent$substrate_type == "percent_fine_substrate", 'perc_total']$perc_total),
         sand_substrate = percent_sand_substrate * (1-substrate_percent[substrate_percent$substrate_type == "percent_sand_substrate", 'perc_total']$perc_total),
         small_gravel = percent_small_gravel_substrate * (1-substrate_percent[substrate_percent$substrate_type == "percent_small_gravel_substrate", 'perc_total']$perc_total),
         large_gravel = percent_large_gravel_substrate * (1-substrate_percent[substrate_percent$substrate_type == "percent_large_gravel_substrate", 'perc_total']$perc_total),
         cobble_substrate = percent_cobble_substrate * (1-substrate_percent[substrate_percent$substrate_type == "percent_cobble_substrate", 'perc_total']$perc_total),
         boulder_substrate = percent_boulder_substrate * (1-substrate_percent[substrate_percent$substrate_type == "percent_boulder_substrate", 'perc_total']$perc_total)) |> 
  select(-c(percent_fine_substrate:percent_boulder_substrate)) |> 
  mutate(woody_debris = sum(percent_large_woody_cover_inchannel, percent_small_woody_cover_inchannel),
         overhead_cover = sum(percent_cover_half_meter_overhead, percent_cover_more_than_half_meter_overhead)
         ) |> 
  select(-c(percent_small_woody_cover_inchannel, percent_large_woody_cover_inchannel, percent_cover_more_than_half_meter_overhead, percent_cover_half_meter_overhead)) |> 
  #filter(species == "Chinook salmon" | is.na(species)) |> 
  select(-count, -channel_geomorphic_unit, -micro_hab_data_tbl_id, -location_table_id, -fish_data_id,
         -focal_velocity, -dist_to_bottom, -fl_mm, -species,  -transect_code, -date, -surface_turbidity) |> 
  select(-(contains("no_cover"))) |> 
  distinct() |> 
  na.omit() |> glimpse()
```

## Logistic Regression: 1 - Using Cover, Substrate, Velocity, and Depth

**Predictors**

-   Depth

-   Velocity

-   Substrate (fine through boulder) normalized by prevalence

-   Woody Debris (`percent_small_woody_cover_inchannel` +
    `percent_large_woody_cover_inchannel`)

-   Overhead Cover (`percent_cover_more_than_half_meter_overhead` +
    `percent_cover_half_meter_overhead`)

-   Submerged Aquatic Vegetation

-   Undercut Bank

**Notes**

-   Chose to remove surface turbidity from predictors because it is not
    a parameter aligned with the strategic plan. This parameter,
    however, was initially found significant in the logistic regression

**Preliminary Results**

-   Significant Predictors include: undercut bank (+), small gravel (-),
    boulder substrate (+), woody debris (+), and overhead cover (+)

#### Build Model

```{r message=FALSE, warning=FALSE}
recipe <- recipe(data = mini_snorkel_grouped, formula = fish_presence ~.) |> 
  # step_poly(all_predictors(), degree = 2) |> x^2
  step_mutate_at(all_numeric_predictors(), fn = asinh) |> # inverse hyperbolic sine as alternative to log that can handle zeros
  step_naomit(all_predictors()) |> # ensure there are no NAs
  step_zv(all_predictors()) |> # ensure there are no columns that are all the same value
  step_normalize(all_numeric_predictors()) # Normalization can facilitate the comparison of the relative importance of different predictors

log_reg <- logistic_reg() |> 
  set_engine("glm") 

log_reg_model <- workflow() |>
  add_recipe(recipe) |>
  add_model(log_reg) |> 
  fit(data = mini_snorkel_grouped)

log_reg_model |> glance()
log_reg_model |> tidy() |> filter(p.value < 0.1)

sig_predictors <- log_reg_model |> tidy() |> filter(p.value < 0.1) |> filter(term != "(Intercept)") |>  pull(term)

predictions <- predict(log_reg_model, mini_snorkel_grouped, type = "prob") |> 
  bind_cols(mini_snorkel_grouped) 

```

```{r eval=FALSE, include=FALSE}
#exploratory to see if we want to change the threshold?
hard_pred_0.5 <- predictions |>
  mutate(.pred = probably::make_two_class_pred(
      estimate = .pred_1,
      levels = levels(fish_presence),
      threshold = .5))  |>
  select(fish_presence, contains(".pred"))

hard_pred_0.5 %>%
  count(.truth = fish_presence, .pred)
```

### Results

The following predictors were found to significantly effect fish
presence: `r paste0(sig_predictors, collapse = ", ")`

```{r message=FALSE, warning=FALSE}
map(sig_predictors, function(predictor) {
  ggplot(predictions, aes_string(x = predictor, y = ".pred_1")) +
    geom_smooth(method = "glm") +
    labs(title = paste("Predicted Probability vs.", predictor),
         y = 'Probability of Fish Presence')
})

```

## Logistic Regression: 2 - Use cover as presence/absence variable

[Beakes et al.
2012](https://www.sfu.ca/biology/faculty/jwmoore/publications/Beakes_etal_RRA_inpress.pdf)
We identified locations with cover when they were within 50 cm of large
woody debris (\>7.5 cm diameter), tall vegetation (\>50 cm above
ground), overhanging vegetation (\<50 cm from water’s surface), large
boulders (\>17.5 cm diameter), undercut banks, large bedrock crevasses
or combinations of these cover types. Areas without cover included
characteristics such as small vegetation, small substrate (\<17.5 cm
diameter) or filamentous algae.

**Predictors**

-   Cover Presence (1/0): summation of undercut bank, woody debris,
    overhead cover, and submerged aquatic vegetation and greater than
    20%

-   Substrate (fine through boulder) normalized by prevalence

-   Depth

-   Velocity

**Preliminary Results**

-   Cover Presence is a significant predictor and the strongest when
    describing probability of fish presence

-   Boulder (+) and Small Grave (-) are also significant predictors

```{r}
mini_snork_cover <- mini_snorkel_grouped |> 
  mutate(cover_total = percent_undercut_bank + woody_debris + overhead_cover + percent_submerged_aquatic_veg_inchannel,
         cover_presence = as.factor(ifelse(cover_total >= 20, 1, 0))) |> 
  select(-c(percent_undercut_bank, woody_debris, overhead_cover, percent_submerged_aquatic_veg_inchannel, cover_total)) |> 
  glimpse()
```

```{r}
recipe <- recipe(data = mini_snork_cover, formula = fish_presence ~.) |> 
  # step_poly(all_predictors(), degree = 2) |> x^2
  step_mutate_at(all_numeric_predictors(), fn = asinh) |> # inverse hyperbolic sine as alternative to log that can handle zeros
  step_naomit(all_predictors()) |> # ensure there are no NAs
  step_zv(all_predictors()) |> # ensure there are no columns that are all the same value
  step_normalize(all_numeric_predictors()) # Normalization can facilitate the comparison of the relative importance of different predictors

log_reg <- logistic_reg() |> 
  set_engine("glm") 

log_reg_model <- workflow() |>
  add_recipe(recipe) |>
  add_model(log_reg) |> 
  fit(data = mini_snork_cover)

log_reg_model |> glance()
log_reg_model |> tidy() |> filter(p.value < 0.1)

sig_predictors <- log_reg_model |> tidy() |> filter(p.value < 0.1) |> filter(term != "(Intercept)") |>  pull(term)

predictions <- predict(log_reg_model, mini_snork_cover, type = "prob") |> 
  bind_cols(mini_snork_cover) 

```

#### Results

The following predictors were found to significantly effect fish
presence: `r paste0(sig_predictors, collapse = ", ")`

```{r}
ggplot(predictions, aes(x = cover_presence, y = .pred_1)) +
    geom_boxplot() +
    labs(title = "Predicted Probability vs. cover presence",
         y = 'Probability of Fish Presence')


map(sig_predictors[1:2], function(predictor) {
  ggplot(predictions, aes_string(x = predictor, y = ".pred_1")) +
    geom_smooth(method = "glm") +
    labs(title = paste("Predicted Probability vs.", predictor),
         y = 'Probability of Fish Presence')
})


```

## Next steps

1.  Compare to species presence/absence
2.  Look into methods for creating a more even sample of fish
    presence/absence that does not create a synthetic dataset
3.  Continue exploring methods for normalization and tuning parameters
    and adding interactions between parameters
4.  Explore using a fixed effect - such as geomorphic unit (pool, glide,
    riffle) or river mile

```{r}
knitr::knit_exit()
```

```{r}
# explore interactions

recipe <- recipe(data = mini_snorkel_grouped, formula = fish_presence ~.) |> 
  step_mutate_at(all_numeric_predictors(), fn = asinh) |> # inverse hyperbolic sine as alternative to log that can handle zeros
  step_naomit(all_predictors()) |> # ensure there are no NAs
  step_zv(all_predictors()) |> # ensure there are no columns that are all the same value
  step_normalize(all_numeric_predictors()) # Normalization can facilitate the comparison of the relative importance of different predictors

log_reg <- logistic_reg() |> 
  set_engine("glm") 

log_reg_model <- workflow() |>
  add_recipe(recipe) |>
  add_model(log_reg) |> 
  fit(data = mini_snorkel_grouped)

interaction_grid <- mini_snorkel_grouped |> 
  select(-fish_presence) |> 
  names() |> 
  combn(2, simplify = FALSE) |> 
  map(~interaction(term = .x[[1]], .x[[2]])) |> 
  unlist()

folds <- vfold_cv(mini_snorkel_grouped, v = 5)
model_with_interactions <- tune_grid(
  log_reg_model,
   resamples = folds,
  grid = interaction_grid
)

show_best(model_with_interactions) 

```

```{r}
# using a single predictor: 
# RESULTS ARE ALIGNED 
recipe <- recipe(data = mini_snorkel_grouped, formula = fish_presence ~ percent_undercut_bank) |> 
  step_mutate_at(all_numeric_predictors(), fn = asinh) |> # inverse hyperbolic sine as alternative to log that can handle zeros
  step_naomit(all_predictors()) |> # ensure there are no NAs
  step_zv(all_predictors()) |> # ensure there are no columns that are all the same value
  step_normalize(all_numeric_predictors()) # Normalization can facilitate the comparison of the relative importance of different predictors

log_reg <- logistic_reg() |> 
  set_engine("glm") 

log_reg_model <- workflow() |>
  add_recipe(recipe) |>
  add_model(log_reg) |> 
  fit(data = mini_snorkel_grouped)

log_reg_model |> tidy()

predictions <- predict(log_reg_model, mini_snorkel_grouped, type = "prob") |> 
  bind_cols(mini_snorkel_grouped) 

ggplot(predictions, aes(x = percent_undercut_bank, y = .pred_1)) +
    geom_smooth(method = "glm") 

```
