---
title: "Mini Snorkel Feather HSI Workflow"
author: "Maddee Rubenson"
date: "`r Sys.Date()`"
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(tidymodels)

knitr::opts_chunk$set(echo = TRUE)

theme_set(theme_minimal())

```

```{r}
# read in mini snorkel data
mini_snorkel <- read_csv('../../feather-mini-snorkel/data/microhabitat_with_fish_observations.csv') |>
  mutate(count = ifelse(is.na(count), 0, count)) |> 
  glimpse()

# mini_snorkel <- read_csv('https://raw.githubusercontent.com/FlowWest/feather-mini-snorkel/qc-data/data/microhabitat_with_fish_observations.csv?token=GHSAT0AAAAAACCSPYJRVFBZ33MHJ2WM5ZX4ZQMRB5A') |> 
#   mutate(count = ifelse(is.na(count), 0, count)) |> 
#   glimpse()

```

## Draft Workflow

[**2004a**](https://netorg629193.sharepoint.com/:b:/s/VA-FeatherRiver/EY9qLwY15ypFn3W42E02takBH745JkefSPAbS4y9VFzuBQ?e=n2QRXK)
Data Analysis

-   Stepwise binary logistic regression analysis was used to assess
    factors influencing the occurrence of steelhead and Chinook salmon.
    Fine scale survey results were analyzed at both the mesohabitat (100
    m2) and microhabitat (1 m2) scale.

-   Mesohabitat analysis was performed by treating the entire 25 m reach
    as a sample

-   Reach habitat variables where steelhead or salmon were present
    (logistic response variable) were compared to reaches where fish
    were absent (logistic reference variable)

-   Microhabitat was done similarly except that individual one square
    meter cells were considered rather than entire reaches

-   Reaches lacking salmon or steelhead were not included in
    microhabitat analysis

[**2005**](https://netorg629193.sharepoint.com/:b:/s/VA-FeatherRiver/ES8H3f7ZUO5Aj7OK8hfW03UB721tCgZd_OEb8P9cbJLiMA?e=L95bf8):
HSC development- chinook salmon rearing

-   HSC were created for fry (\<50 mm) and juvenile Chinook salmon depth
    and mean column velocity data using a three-point running mean to
    smooth frequency distributions of the fry habitat use data and using
    NPTL [what does this stand for?] for the juvenile habitat use data.

-   all substrate suitability given value of 1 after finding that
    substrate was not driving factor for microhabitat selection of fry
    and juveniles

-   cover modified to: with and without. Suitability of without cover
    was calculated as the percentage of fish observed without cover to
    the total sample size

-   suitability of cover present was assigned a value of 1.0 and 0.30 or
    0.22 for cover absent for Chinook salon fry and juveniles [I do not
    get this... ]

-   Focus on suitability of instream cover. When a cover variable did
    not exist, preference for low velocity and shallow depth in a large
    river indicate suitable habitat along stream margins or out in the
    main channel when the river is nearly dry and the preferred
    conditions are prevalent.

-   intermediate scale data separated into four cover types: no cover,
    object only cover overhead only cover, both object and overhead
    cover

[**Gard
2023**](https://netorg629193.sharepoint.com/:b:/s/VA-FeatherRiver/EelfImRfhzxKjLQdrstbIPgBqbRV5S0Wke5GkSh06_CUrQ?e=6PdWKh) -
HSC comparison

-   Presence/absence HSC are developed using a polynomial logistic
    regression that uses both the occupied and unoccupied data; the
    results are then rescaled that the highest value is 1 to calculate
    the HSI

-   For depth and velocity HSC, the criteria were developed directly
    from use observations using a range of curve fitting and smoothing
    techniques

-   Use/availability criteria are developed by dividing use
    observations, generally binned, by availability data from transects

```{r}

mini_snorkel_model_ready <- mini_snorkel |> 
  select(depth:percent_cover_more_than_half_meter_overhead, count, channel_geomorphic_unit, species) |> 
  mutate(fish_presence = as.factor(ifelse(count < 1, "0", "1")))

```

# Chinook Salmon

### Logistic Regression

Build a logistic regression of fish presence and absence to identify
important cover and substrate variables.

**Findings**

-   Substrate: most important predictor is small gravel

**Questions**

-   Substrate: boulder came out as NA in model. Should explore why this
    occurred.

```{r}
# start with Chinook and substrate 
chn_mini_snorkel <- mini_snorkel_model_ready |> 
  filter(species == "Chinook salmon" | is.na(species)) |> 
  select(fish_presence, percent_fine_substrate:percent_boulder_substrate)

# Define a recipe
rec <- recipe(fish_presence ~  percent_fine_substrate + percent_sand_substrate + percent_small_gravel_substrate + percent_large_gravel_substrate +
              percent_cobble_substrate + percent_boulder_substrate, data = chn_mini_snorkel)  

# Split the data into training and testing sets
data_split <- initial_split(chn_mini_snorkel, prop = 0.8, strata = "fish_presence")
data_train <- training(data_split)
data_test <- testing(data_split)

# Create a logistic regression model
log_reg <- logistic_reg() |> 
  set_engine("glm") |> 
  set_mode("classification") |> translate()

# Create a workflow
wf <- workflow()  |> 
  add_recipe(rec) |> 
  add_model(log_reg)

# Train the model
wf_fit <- wf |> 
  fit(data_train)

tidy(wf_fit) # why is boulder NA? small gravel only significant predictor

# Make predictions
# predictions <- predict(wf_fit, data_test) |> 
#   bind_cols(data_test)
```

## Adapted Mark Gard 1998 HSC process to Cover and Substrate

Mark Gard's 1998 paper identified a method to developing HSC for
presence and absence of redds in different substrate classes. I will
follow the same methodology using fish presence/absence instead of redds
and also apply to cover.

1.  Determine number of fish with each substrate-size class
2.  Calculate the proportion of fish with each substrate size class
3.  Calculate the HSC value for each substrate size class by dividing
    the proportion of fish in each substrate class by the proportion of
    fish with the most frequent substrate class

To convert from percent to discrete:

-   Cover decision: 20% or greater is considered present

-   Substrate decision: used same criteria as cover (for now)

**Discussion**

Substrate:
-   Sand has a stronger HSI when looking at number of fish vs.
    presence/absence

-   Presence/absence results are consistent with logistic regression

Cover:
-   More consistent results when looking at number of fish vs. presence/absence 

```{r}
mark_gard_1998_hsi <- function(data, percent_presence = 20, cols = c()) {
  data_tidy <- data |> 
    select(count, fish_presence, cols) |> 
    pivot_longer(cols = cols, names_to = "type", values_to = "percent") |> 
    mutate(presence = ifelse(percent >= percent_presence, 1, 0)) |> 
    filter(presence == 1) |> 
    group_by(type) |>
    summarise(n_fish_presence = sum(as.numeric(fish_presence)),
              n_fish_total = sum(count)) |> 
    mutate(prop_fish_presence = n_fish_presence/sum(n_fish_presence),
           prop_fish_total = n_fish_total/sum(n_fish_total)) |> 
    ungroup()
  
     most_freq_total <- data_tidy |> 
      arrange(desc(prop_fish_total)) |> 
      slice(1) 
 
     most_freq_presence <- data_tidy |> 
      arrange(desc(prop_fish_presence)) |> 
      slice(1) 
  
  
  data_tidy_final <- data_tidy |> 
    mutate(hsc_presence = prop_fish_presence/most_freq_presence$prop_fish_presence,
           hsc_total = prop_fish_total/most_freq_total$prop_fish_total)
  
  return(data_tidy_final)

}
```

### Substrate

```{r}
cols <- c('percent_fine_substrate', "percent_sand_substrate" , 'percent_small_gravel_substrate', 'percent_large_gravel_substrate', 'percent_cobble_substrate', 'percent_boulder_substrate')

hsi_substrate <- mark_gard_1998_hsi(data = mini_snorkel_model_ready, percent_presence = 20, cols = cols)

hsi_substrate |> 
  pivot_longer(cols = c(hsc_presence, hsc_total), names_to = "hsc", values_to = "values") |> 
  ggplot() + 
  geom_col(aes(y = values, x = type, fill = hsc), position = "dodge") +
  coord_flip() +
  ggtitle('Mark Gard 1998: Substrate HSI Comparison')

```

### Cover

```{r}
cols <- c('percent_no_cover_inchannel', 'percent_small_woody_cover_inchannel', 'percent_large_woody_cover_inchannel', 
          'percent_submerged_aquatic_veg_inchannel', 'percent_undercut_bank', 'percent_no_cover_overhead', 'percent_cover_half_meter_overhead',
          'percent_cover_more_than_half_meter_overhead')

hsi_cover <- mark_gard_1998_hsi(data = mini_snorkel_model_ready, percent_presence = 20, cols = cols)

hsi_cover |> 
  pivot_longer(cols = c(hsc_presence, hsc_total), names_to = "hsc", values_to = "values") |> 
  ggplot() + 
  geom_col(aes(y = values, x = type, fill = hsc), position = "dodge") +
  coord_flip() +
  ggtitle('Mark Gard 1998: Cover HSI Comparison')
```

```{r}
knitr::knit_exit()
```

### Mapping of Substrate and Cover

| Mini Snorkel Data Type | Substrate Size Range (mm) | Gard 2023 Code |
|------------------------|---------------------------|----------------|
| fine                   | 0-0.05                    | 0.1            |
| sand                   | 0.05-2                    | 0.1            |
| small gravel           | 2-50                      | 1 - 1.2        |
| large gravel           | 50-150                    | 1.3-4.6        |
| cobble                 | 150-300                   | 6.8 - 8        |
| boulder                | \>300                     | 9              |

This mapping is a work in progress. The codes don't make a lot of sense
and there is uncertainty on how these codes were developed. A next step
would be to discuss the 2023 paper with Gard to get a better
understanding of how the codes were used in his workflow.
