---
title: "Feather River - Mini Snorkel Cover Analysis"
output: 
  flexdashboard::flex_dashboard:
    theme: readable
  font_family: "Roboto"
  google_fonts: true
runtime: shiny
---

```{r global, include=FALSE}
library(tidyverse)
library(leaflet)
library(sf)

colors_full <-  c("#9A8822", "#F5CDB4", "#F8AFA8", "#FDDDA0", "#74A089", #Royal 2
                  "#899DA4", "#C93312", "#DC863B", # royal 1 (- 3)
                  "#F1BB7B", "#FD6467", "#5B1A18", "#D67236",# Grand Budapest 1 (-4)
                  "#D8B70A", "#02401B", "#A2A475", # Cavalcanti 1
                  "#E6A0C4", "#C6CDF7", "#D8A499", "#7294D4", #Grand Budapest 2
                  "#9986A5", "#EAD3BF", "#AA9486", "#B6854D", "#798E87", # Isle of dogs 2 altered slightly
                  "#F3DF6C", "#CEAB07", "#D5D5D3", "#24281A", # Moonriese 1, 
                  "#798E87", "#C27D38", "#CCC591", "#29211F", # moonrise 2
                  "#85D4E3", "#F4B5BD", "#9C964A", "#CDC08C", "#FAD77B" ) # moonrise 3 

fish_pres_cols <- c('Yes' = "#9A8822", 'No' = "#899DA4")

colfunc <- colorRampPalette(c("#02401B", "white"))
gradient <- colfunc(100)

```

```{r}
mini_locations_raw <- read_csv(here::here("visualizations", "mini_locations_raw.csv"))
mini_fish_raw <- read_csv(here::here("visualizations","mini_fish_raw.csv"))
# orders locations from upstream to downstream. 
# TODO we are missing coordinates for some locations
location_order <- tibble(location = c("hatchery ditch", "hour bars", "hatchery riffle", "trailer park",
                                      "bedrock riffle", "steep riffle", "robinson riffle", "upper big hole",
                                      "lower big hole", "g95", "gridley riffle", "big bar", "goose riffle",
                                      "macfarland riffle", "shallow riffle", "herringer riffle", "junkyard riffle",
                                      "auditorium riffle", "eye side channel", "matthews riffle", "eye riffle",
                                      "lower hour", "aleck riffle", "weir riffle", "vance avenue",
                                      "big hole", "hour riffle", "cox riffle", "gateway riffle"),
                         order = c(2,19,1,5,
                                   4,9,8,15,
                                   16,17,23,21,20,
                                   22,25,26,24,
                                   3,11,6,12,
                                   18,7,10,13,
                                   14,17,25,13))

steelhead_redds <- read_rds(here::here("visualizations","steelhead_redd_summary.rds")) |> 
  mutate(species = "steelhead trout (clipped)") 
chinook_redds <- read_rds(here::here("visualizations","chinook_redd_summary.rds")) |> 
  mutate(species = "chinook salmon") 

all_redds <- bind_rows(steelhead_redds, chinook_redds)

feather_cover <- mini_fish_raw |> 
  left_join(mini_locations_raw |> 
              distinct(location_table_id, date, channel_location, location)) |> 
  # create cover variables that are more consistent with strategic plan categories
  mutate(small_woody_debris = percent_small_woody_cover_inchannel,
         large_woody_debris = percent_large_woody_cover_inchannel,
         boulder = percent_boulder_substrate,
         cobble = percent_cobble_substrate, 
         undercut_bank = percent_undercut_bank, 
         aquatic_veg = percent_submerged_aquatic_veg_inchannel, 
         overhanging_veg = percent_cover_half_meter_overhead + percent_cover_more_than_half_meter_overhead) |> 
  select(micro_hab_data_tbl_id, location_table_id, transect_code, location,
         small_woody_debris:overhanging_veg, count, date, depth, velocity, species) |>
  left_join(all_redds, by = c("location", "species"))  |> 
  pivot_longer(cols = c(small_woody_debris, boulder, cobble, undercut_bank, aquatic_veg, overhanging_veg, depth, velocity, redd_presence, redd_total), names_to = "cover_type", 
               values_to = "percent_cover") |>  
  distinct() |> 
  left_join(location_order) |> 
  mutate(fish_presence = ifelse(count > 0, "Yes", "No")) |> select(-count) |> 
  mutate(location_table_id = paste0(date, " (location id: ", location_table_id, ")")) |> 
  filter(species %in% c("chinook salmon" , "steelhead trout (wild)", "steelhead trout, (clipped)")) 

summary_cover <- feather_cover |> 
  mutate(month = lubridate::month(date, label = TRUE)) |> 
  group_by(location, cover_type, order, fish_presence, month, species) |> 
  summarize(
    mean_percent_cover = if (first(cover_type) %in% c("redd_presence", "redd_total")) {
      max(percent_cover, na.rm = TRUE)
    } else {
      mean(percent_cover, na.rm = TRUE)
    },
    median_percent_cover = if (first(cover_type) %in% c("redd_presence", "redd_total")) {
      max(percent_cover, na.rm = TRUE)
    } else {
      median(percent_cover, na.rm = TRUE)
    },
    min_percent_cover = min(percent_cover, na.rm = TRUE),
    max_percent_cover = max(percent_cover, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  mutate(mean_percent_cover = ifelse(is.na(mean_percent_cover), 0, mean_percent_cover))

location_summary <- feather_cover |> 
  group_by(location, cover_type, species) |> 
  summarize(mean_percent_cover = mean(percent_cover, na.rm = T)) |> 
  mutate(mean_percent_cover = ifelse(is.na(mean_percent_cover), 0, mean_percent_cover))

locations_sf <- mini_locations_raw |>
  distinct(location_table_id, date, channel_location, .keep_all = TRUE) |>
  na.omit() |>
  st_as_sf(coords = c('longitude', 'latitude'), crs = 4326) |> 
  group_by(location) |>
  slice_max(geometry) |>
  distinct() |> 
  select(location) |> 
  left_join(location_summary)


```

```{r}
imp_steelhead_pred <- c('redd_presence', 'overhanging_veg', 'aquatic_veg', 'small_woody_debris', 'redd_total', 'depth', 'velocity')

imp_chinook_pred <- c("overhanging_veg", "depth", "redd_total", "velocity")
```

# About the Tool 

#### About the Tool 

The purpose of this tool is to show the spatial and temporal variability in cover within the Feather River system using 2001 and 2002 mini snorkel data. For both species, only important predictors are options to view based on logistic regression results from a prior analysis.

TODO: link to EDI

TODO: link to prior app

TODO: link to analyses

#### Habitat Map

This map shows all sampling locations, however note that there are mutliple transect locations within each sampling location point that are not represented. Color and size represent the average of selected predictor.

#### Cover Plot

Plot represents the temporal and cover variability of each predictor. The bars range from minimum to maximum and the points are the mean. Values are summarized by each location over all transects.

# View Analysis Inputs {#view-data data-orientation="rows"}

## Row {data-height="25"}

```{r}
# Add the radio buttons with a help icon
tags$div(
  tags$strong("Choose which model you want to view:"),
  tags$span(
    class = "help-icon", 
    shiny::icon("question-circle"), 
    style = "cursor: pointer; color: #007BFF; margin-left: 5px;",
    onclick = "document.getElementById('instructions_modal').style.display='block'; 
               document.getElementById('modal_overlay').style.display='block';"
  ),
  radioButtons(
    inputId = "which_fish", 
    label = NULL, 
    choices = c("Steelhead", "Chinook"), 
    selected = "Steelhead"
  )
)

# Hidden modal for instructions
tags$div(
  id = "modal_overlay",
  style = "display:none; position:fixed; top:0; left:0; width:100%; height:100%; background-color:rgba(0, 0, 0, 0.5); z-index:999;",
  onclick = "document.getElementById('instructions_modal').style.display='none'; 
             document.getElementById('modal_overlay').style.display='none';"
)

tags$div(
  id = "instructions_modal",
  style = "display:none; position:fixed; z-index:1000; left:50%; top:50%; transform:translate(-50%, -50%); background-color:white; padding:20px; border:1px solid #ccc; box-shadow:0px 4px 8px rgba(0,0,0,0.2);",
  tags$div(
    style = "text-align:right;",
    tags$button("Close", 
                onclick = "document.getElementById('instructions_modal').style.display='none'; 
                           document.getElementById('modal_overlay').style.display='none';", 
                style = "background:none; border:none; color:#007BFF; font-weight:bold; cursor:pointer;")
  ),
  tags$h4("Instructions"),
  tags$p("Attribute Filter - Select a monitoring location by name on the drop down. This will populate which transects were monitored at that monitoring location. Each transect is associated with a specific monitoring date and ID. Once selected, a heat map of percent cover along the transect will be visible. If a fish was present during monitoring, it will be highlighted with a red outline.")
)


```

```{r}
# attribute view conditional panel 
conditionalPanel(
  condition = "input.which_fish == 'Steelhead'",
  div(
    style = "display: flex; align-items: flex-start; gap: 10px;", # Flexbox for horizontal layout
    selectInput(
      "location_filter",
      label = tags$strong("Filter by Location:"),
      choices = c("all locations", unique(locations_sf$location)),
      selected = "all locations"
    ),
    output$predictor_id_ui <- renderUI({
      selectInput("predictor_id", label = tags$strong("Choose Predictor:"),
                  choices = c(imp_steelhead_pred), selected = "aquatic_veg")
    })
  )
)

```

```{r}
# map view conditional panel 
conditionalPanel(
  condition = "input.which_fish == 'Chinook'",
  div(
    style = "display: flex; align-items: flex-start; gap: 10px;", # Flexbox for horizontal layout
    selectInput(
      "location_filter",
      label = tags$strong("Filter by Location:"),
      choices = c("all locations", unique(locations_sf$location)),
      selected = "all locations"
    ),
    output$predictor_id_ui <- renderUI({
      selectInput("predictor_id", label = tags$strong("Choose Predictor:"),
                  choices = c(imp_chinook_pred), selected = "aquatic_veg")
    })
  )
)

```

## Row

### Habitat Map

```{r}
output$map <- renderLeaflet({
  
  req(input$predictor_id)
  req(input$which_fish)
  
  map_data <- locations_sf |> filter(cover_type == input$predictor_id) |> 
    filter(
      species %in% if (input$which_fish == "Steelhead") {
        c("steelhead trout (wild)", "steelhead trout (clipped)")
      } else {
        "chinook salmon"
      }
    )
  
  pal <- colorBin("YlOrRd",  map_data$mean_percent_cover, bins = 5)
  
  leaflet(map_data) |>
    addProviderTiles(providers$Esri.WorldImagery, group = "Aerial Imagery") |>
    addProviderTiles(providers$OpenStreetMap, group = "Street Map") |>
    addCircleMarkers(
      layerId = ~location,
      color = "darkblue",
      fillColor = ~pal(mean_percent_cover),
      radius = ~scales::rescale(mean_percent_cover, to = c(3, 20)),
      fill = TRUE,
      fillOpacity = 0.9,
      opacity = 0.5,
      label = ~location,
      labelOptions = labelOptions(
        style = list("font-weight" = "bold", padding = "3px 8px"),
        textsize = "10px",
        direction = "auto",
        noHide = F
      )
    ) |> 
    addLayersControl(
      baseGroups = c("Aerial Imagery", "Street Map"),
      position = "topleft",
      options = layersControlOptions(collapsed = FALSE)
    ) |> 
    addLegend("bottomright",
              pal = pal,
              values = map_data$mean_percent_cover,
              title = input$predictor_id)
})

leafletOutput("map")

observeEvent(input$location_filter, {
  req(input$location_filter)
  
  if (input$location_filter != "all locations") {
    
    selected_location <- locations_sf |>
      ungroup() |>
      filter(location == input$location_filter)
    
    req(nrow(selected_location) > 0)
    
    coords <- sf::st_coordinates(selected_location)
    lng <- as.numeric(coords[1, "X"])
    lat <- as.numeric(coords[1, "Y"])
    
    leafletProxy("map") |>
      flyTo(
        lng = lng,
        lat = lat,
        zoom = 16
      ) |>
      clearGroup("highlighted") |>
      addCircleMarkers(
        data = selected_location,
        radius = 8,
        color = "red",
        weight = 4,
        fill = TRUE,
        fillOpacity = 0.25,
        opacity = 1,
        label = ~location,
        labelOptions = labelOptions(
          noHide = TRUE,
          direction = "auto",
          textsize = "10px",
          style = list("font-weight" = "bold", padding = "3px 8px")
        ),
        group = "highlighted"
      )
  } else {
    leafletProxy("map") |>
      clearGroup("highlighted")
  }
})


```

### Cover Visual

```{r}
output$attribute_plot <- renderPlot({
  if (input$location_filter != "all locations") {
    
    summary_cover_filter <- summary_cover |> 
      filter(location == input$location_filter) |> 
      filter(cover_type == input$predictor_id) |> 
      filter(
        species %in% if (input$which_fish == "Steelhead") {
          c("steelhead trout (wild)", "steelhead trout (clipped)")
        } else {
          "chinook salmon"
        }
      )
    
    ggplot(summary_cover_filter,
           aes(x = month,
               y = mean_percent_cover,
               color = fish_presence,
               group = fish_presence)) +
      geom_linerange(
        aes(ymin = min_percent_cover,
            ymax = max_percent_cover),
        position = position_dodge(width = 0.5),
        size = 2,
        alpha = 0.4
      ) +
      geom_point(
        position = position_dodge(width = 0.5),
        size = 4
      ) +
      scale_color_manual(values = fish_pres_cols) +
      theme_minimal() +
      labs(
        y = "% Cover",
        x = "Month",
        color = "Fish Presence"
      ) +
      theme(
        axis.text = element_text(size = 16),
        axis.title = element_text(size = 18, face = "bold"),
        axis.ticks = element_line(size = 1),
        axis.ticks.length = unit(0.25, "cm"),
        legend.title = element_text(size = 18, face = "bold"),
        legend.text = element_text(size = 16)
      )
    
  } else {
    
    summary_cover_filter <- summary_cover |> 
      filter(cover_type == input$predictor_id) |> 
      filter(
        species %in% if (input$which_fish == "Steelhead") {
          c("steelhead trout (wild)", "steelhead trout (clipped)")
        } else {
          "chinook salmon"
        }
      )
    
    
    ggplot(summary_cover_filter,
           aes(x = month, y = reorder(location, order, decreasing = T), fill = mean_percent_cover)) +
      geom_tile(alpha = 0.8, color = "darkgrey") +
      scale_fill_viridis_c(option = "mako") + 
      labs(y = "River location", x = "Month", fill = input$predictor_id) + 
      theme_minimal() + 
      theme(
        axis.text = element_text(size = 16),
        axis.title = element_text(size = 18, face = "bold"),
        axis.ticks = element_line(size = 1),
        axis.ticks.length = unit(0.25, "cm"),
        legend.title = element_text(size = 18, face = "bold"),
        legend.text = element_text(size = 16)
      ) 
  }
  
  
})

```

```{r}
plotOutput("attribute_plot")
```
