---
title: "Feather River - Mini Snorkel Data Modeling and Analysis"
author: "Maddee Wiggins (FlowWest)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(sjPlot)

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)

theme_set(
  theme_minimal() +
  theme(
    plot.title = element_text(size = 11, face = "bold"),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14)
  )
)


custom_colors <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

options(scipen=999)

```

## Objective

Develop a model that reflects the significance of cover, substrate,
depth, and velocity on fish presence and absence.

## Methodology

**TODO: this section must be updated**

Model was developed using [Feather River Mini Snorkel
Data](https://github.com/FlowWest/feather-mini-snorkel). The data is
structured with a numeric fish count variable that can be converted to a
binary presence/absence. Logistic regression models were tested using
the presence/absence as the response and cover, substrate, depth and
velocity as the explanatory variables. However, due to the skewness of
absence values, none of the developed models were well fit to the data.
Based on a Gard 2024 paper (in-review), we decided to explore the use of
hurdle models. Hurdle models excel with data types that are heavily
skewed towards absence.

```{r}
source(here::here("data-raw", "pull_from_edi.R"))

mini_snorkel_raw <- mini_fish_raw |> 
  left_join(mini_locations_raw |> distinct()) |> 
  mutate(count = ifelse(is.na(count), 0, count),
        fish_presence = as.factor(ifelse(count < 1, "0", "1")),
        month = month(date)) |> 
  glimpse()

```

### Review Data

#### Outliers

`count` outliers exist, in the high flow and the low
flow channel, however their removal did not impact the model
results so they were kept in the dataset.

```{r eval=FALSE, include=FALSE}
# Keeping this figure out of analysis but keeping for future
mini_snorkel_raw |>
  ggplot() +
  geom_boxplot(aes(x = channel_location, y = count)) +
  geom_jitter(aes(x = channel_location, y = count), alpha = 0.5) +
  ggtitle("Outliers of fish count")
```

We thought that by removing values greater than 250 this would limit overdispersion
in the count model, however, it did not. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
mini_snorkel_raw |> 
  filter(count > 0) |> 
  ggplot() + 
  geom_histogram(aes(count, fill = channel_location), binwidth = 50) + 
  scale_fill_manual(values = custom_colors) + 
  ylab("fish count") +
  ggtitle('Figure 1. Histogram of fish count in the high and low flow channels') +facet_grid(~channel_location)

mini_snorkel_model_ready <- mini_snorkel_raw
```

### High Flow vs. Low Flow Channel

Table 1 and figure 2 explore whether fish presence was impacted by the
high or low flow channels. Overall there are similar numbers of 
fish present in the high flow channel compared to the low flow channel
(table 1). There are more fish present in the high flow channel in March but they
move quickly downstream. Fish remain in the low flow channel for much longer time (figure 2).

do we have the same amount of sampling plots in the high flow vs. low flow? 

```{r echo=FALSE, message=FALSE, warning=FALSE}

# looking at overall count data between the low flow/high flow channels
mini_snorkel_model_ready |> 
  group_by(channel_location) |> 
  tally(count) |> 
  knitr::kable(caption = "Table 1. Total count of fish between high flow and low flow channels")

mini_snorkel_model_ready |> 
  group_by(channel_location) |> 
  summarize(n_sites = length(unique(location))) |> 
  knitr::kable(caption = "Table 2. Number of sampling sites in the high flow and low flow channels")

# This figure shows how there are more fish present in the high flow channel 
# for less time. The fish stick around for much longer in the low flow channel. 
mini_snorkel_model_ready |> 
  group_by(channel_location, month = lubridate::month(date)) |> 
  tally(count) |> 
  ggplot() + 
  geom_col(aes(x = as.factor(month), y = n, fill = channel_location), position = "dodge") + 
  ylab("fish count") +
  xlab("month") +
  ggtitle('Figure 2. Fish presence as a function of month and high flow/low flow channel') + 
  scale_fill_manual(values = custom_colors)
  
```

### Variables of Interest

The variables of interest include cover, substrate, velocity and depth variables
known to be important for salmon rearing habitat.

- Velocity - numeric
- Depth - numeric

Substrate and cover variables are measured as percentages which can also be modeled
as binary by establishing a threshold percentage. Overhanging vegetation was measured
at 1/2 meter overhead and more than 1/2 meter overhead. These categories were combined
for simplicity and for comparison with other studies, such as those by Mark Gard.

- Undercut bank
- Aquatic vegetation
- Overhanging vegetation
- Small woody cover
- Large woody cover
- Boulder substrate
- Cobble substrate

Data was collected between March through August, however, outmigration affects
fish presence later in the season. We do not include month as it is a strong predictor
of fish presence. The analysis could be limited to focus on the month with most
observations (March), though including all months also makes sense if seeking to 
determine variables that are stable throughout time.

The conditions in the HFC and LFC are quite different and some initial exploration
was done to explore this by creating separate models for the LFC and HFC. The results
did not differ significantly. Spatial nuances may best be explored at the site level.

## Hurdle Model Approach

### Hurdle Models and Interpretation

A hurdle model was used in Gard 2024 (in-review) to test for the effects
of cover and habitat type on the total abundance of Chinook salmon at
both site and cell level. Here we explore the use oif a hurdle model to help
understand the influence of velocity, depth, and cover on fish count and
presence/absence.

**Hurdle Models**

Hurdle models are used when count data has an excess of zeros. These
models can be understood as a mixture of two subset of populations. In
one subset, we have a usual count model that may or may not generate
zero, and the other subset only produce zero count.

A hurdle model models excess zeroes separately from the rest of the
data. The zero counts are modeled as a binary response variable and the
positive counts are modeled using poisson distribution.

*Interpreting a Hurdle Model*

The binary part of the model helps identify factors that influence the
presence/absence of fish. The coefficients of the zero part of the
hurdle model represent the odds ratio of observing at least one fish.

The count part of the model estimate the effects of predictor variables
on the count outcome, excluding all zero counts. Coefficients of counts
represent rate ratios of one or more fish observed.

The Incidence Result Ratio (IRR) in the count part of the model (count
\> 0) represent the multiplicative effect of a one-unit change in a
predictor variable on the expected count of non-zero observations,
assuming all other variables are held constant. For example, if the IRR
for a predictor is 1.2, it means that a one-unit increase in that
predictor is associated with a 20% increase in the expected count of
non-zero observations, assuming all other variables remain constant. For
the binary part of the model - if the coefficient for a predictor in the
binary part of the hurdle model is 0.5, it means that a one-unit
increase in the predictor is associated with a 50% increase in the odds
of having a zero count versus a positive count, assuming all other
variables are held constant.

<!-- #### The Binary Predictor Model -->

<!-- This model utilizes binary cover and substrate variables. The results do not differ -->
<!-- that much from using the absolute percentage variables for cover and substrate. -->

<!-- - The count model has extremely high overdispersion. This is not surprising as -->
<!-- most of the count variables are 1 with few observations greater than 1. -->
<!-- - The intercept of the zero model is significant. This is not surprising as the  -->
<!-- number of zeros is nearly 96%. -->

### Build Model 

All cover variables were converted to presence/absence using a threshold of 20%. The following is the data structure of the model input data. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
percent_threshold <- 20

# use a hurdle model (Gard 2024); 
# pre-processing - all numeric values must be rounded to whole numbers
hurdle_data <- mini_snorkel_model_ready |>
    select(count, location, channel_location, depth, velocity, contains("inchannel"), contains("overhead"), "percent_cobble_substrate", "percent_boulder_substrate", "percent_undercut_bank", month, fl_mm) |>
  # create new cover variables, based off Gard 2023
  mutate(small_woody = percent_small_woody_cover_inchannel,
        large_woody = percent_large_woody_cover_inchannel,
         boulder_substrate = percent_boulder_substrate,
         cobble_substrate = percent_cobble_substrate, 
         undercut_bank = percent_undercut_bank, 
         aquatic_veg = percent_submerged_aquatic_veg_inchannel, 
         overhanging_veg = percent_cover_half_meter_overhead + percent_cover_more_than_half_meter_overhead) |>
  # transform to presence/absence based on a defined threshold
  mutate(cobble_substrate = ifelse(percent_cobble_substrate >= percent_threshold, 1, 0 ),
         boulder_substrate = ifelse(percent_boulder_substrate >= percent_threshold, 1, 0 ),
         small_woody = ifelse(small_woody >= percent_threshold, 1, 0 ),
         large_woody = ifelse(large_woody >= percent_threshold, 1, 0 ),
         aquatic_veg = ifelse(aquatic_veg >= percent_threshold, 1, 0),
         undercut_bank = ifelse(undercut_bank >= percent_threshold, 1, 0),
         no_cover_overhead = ifelse(percent_no_cover_overhead >= percent_threshold, 1, 0),
         overhanging_veg = ifelse(overhanging_veg >= percent_threshold, 1, 0)) |>
  #remove no cover variables (no cover overhead, no cover in channel)
  select(-contains("no_cover")) |> 
  distinct() |> 
  mutate(fl_mm = ifelse(is.na(fl_mm), 0, fl_mm)) |> 
  na.omit() |> 
  select(-fl_mm) |> 
  glimpse()

```

### Results summary

```{r}
hurdle_model <- pscl::hurdle(count ~ small_woody + depth + velocity + large_woody + aquatic_veg + overhanging_veg + cobble_substrate + boulder_substrate + undercut_bank, data = hurdle_data, dist = "negbin") 
# Negative Binomial Distribution: The Negative Binomial distribution is more flexible than the Poisson distribution and is used when the variance of the counts is greater than the mean, which is known as overdispersion. The Negative Binomial model allows for the variance to be larger than the mean, which is common in count data where there is extra variability beyond what is expected from a Poisson distribution.

# zero inflated models are capable of dealing with excess zero counts
# NOTE: Chose to not use a zero inflated model because it operates under the assumption that excess zeros are generated by a separate process from the count values. 
# zeroinf_model <- pscl::zeroinfl(count ~ ., data = hurdle_data) # , zero.dist = "binomial", dist = "negbin"

best_model_hurdle <- MASS::stepAIC(hurdle_model)
summary(best_model_hurdle)
```


```{r}
plot_model(best_model_hurdle, title = "Figure 3. Hurdle Model Results Using Binary Presence/Absence for Cover and Substrate",  wrap.title = 50) + 
  scale_color_manual(values = custom_colors[5:6])

```

```{r}
s <- summary(best_model_hurdle)

# Extract components
count_coef <- as.data.frame(s$coefficients$count) |>
  tibble::rownames_to_column("term") |>
  rename(Estimate = Estimate,
         Std.Error = `Std. Error`,
         z = `z value`,
         p.value = `Pr(>|z|)`) |>
  mutate(
    effect_type = "Abundance (Count Multiplier)",
    exp_effect = exp(Estimate)
  )

zero_coef <- as.data.frame(s$coefficients$zero) |>
  tibble::rownames_to_column("term") |>
  rename(Estimate = Estimate,
         Std.Error = `Std. Error`,
         z = `z value`,
         p.value = `Pr(>|z|)`) |>
  mutate(
    effect_type = "Presence (Odds Ratio)",
    exp_effect = exp(Estimate)
  )

# Combine & clean
effect_table <- bind_rows(zero_coef, count_coef) |>
  mutate(
    term = recode(term,
      "(Intercept)" = "Intercept",
      small_woody = "Small woody debris",
      large_woody = "Large woody debris",
      overhanging_veg = "Overhanging vegetation",
      aquatic_veg = "Aquatic vegetation",
      cobble_substrate = "Cobble substrate",
      boulder_substrate = "Boulder substrate",
      undercut_bank = "Undercut bank",
      depth = "Depth",
      velocity = "Velocity"
    ),
    exp_effect = round(exp_effect, 2),
    Estimate   = round(Estimate, 3),
    Std.Error  = round(Std.Error, 3),
    p.value    = signif(p.value, 3)
  ) |>
  rename(
    Variable = term,
    `Effect type` = effect_type,
    `Exp(effect)` = exp_effect
  )

knitr::kable(effect_table, align = "lcccc")

```

### Model Performance 

#### Evaluate the presence / absence component (classification)
```{r}
pred_presence <- predict(
  best_model_hurdle,
  type = "response",
  which = "zero"
)
obs_presence <- as.integer(hurdle_data$count > 0)

library(pROC)
roc_presence <- roc(obs_presence, pred_presence)
auc(roc_presence)

# Interpretation:
# AUC = 0.5 → no better than random
# 0.7–0.8 → acceptable
# 0.8–0.9 → good
# >0.9 → excellent

pred_class <- ifelse(pred_presence > 0.5, 1, 0)
table(Predicted = pred_class, Observed = obs_presence)

```

#### Evaluate the count component (abundance, conditional on presence)

Only evaluate sites where count > 0.

```{r}
present_idx <- hurdle_data$count > 0

pred_count <- predict(
  best_model_hurdle,
  type = "response",
  which = "count"
)[present_idx]

obs_count <- hurdle_data$count[present_idx]

rmse  <- sqrt(mean((obs_count - pred_count)^2))
mae   <- mean(abs(obs_count - pred_count))
r2    <- cor(obs_count, pred_count)^2
c(rmse = rmse, mae = mae, r2 = r2)

# Interpretation:
# Lower RMSE/MAE = better fit
# Higher R² = stronger correspondence

plot(pred_count, obs_count)
abline(0,1,col="red")

```

#### Evaluate the full “combined” prediction

```{r}
full_pred <- pred_presence * predict(
  best_model_hurdle,
  type = "response",
  which = "count"
)

obs <- hurdle_data$count

rmse_full <- sqrt(mean((obs - full_pred)^2))
mae_full  <- mean(abs(obs - full_pred))
r2_full   <- cor(obs, full_pred)^2
c(rmse_full, mae_full, r2_full)


```

### Hurdle Model Discussion

The hurdle model is performing okay (AUC of `r auc(roc_presence)`) for presence/absence but there is a poor predictive skill for the count variable likely due to many counts being low but with high variability. Because of this, we chose to model habitat associations for fish occurrence using a logistic model, which is most consistent with the information content of the data. 

```{r}
hist(hurdle_data$count[hurdle_data$count>0])
```

## Logistic Regression Approach 

```{r}
log_reg_data <- hurdle_data |> 
  mutate(presence = as.integer(count > 0))
```

### Build Logistic Regression Model 

```{r}
glm_presence <- glm(
  presence ~ small_woody + depth + velocity + large_woody +
    aquatic_veg + overhanging_veg + cobble_substrate +
    boulder_substrate + undercut_bank,
  data = log_reg_data,
  family = binomial(link = "logit")
)

summary(glm_presence)

```

**Odds Ratio**


```{r}
presence_table <- broom::tidy(glm_presence, conf.int = TRUE) |>
  mutate(
    odds_ratio = exp(estimate),
    or_low = exp(conf.low),
    or_high = exp(conf.high)
  ) |>
  transmute(
    term,
    estimate = round(estimate, 3),
    `Std.Error` = round(std.error, 3),
    `Odds ratio` = round(odds_ratio, 2),
    `95% CI` = sprintf("[%.2f, %.2f]", or_low, or_high),
    `p` = signif(p.value, 3)
  )

knitr::kable(presence_table)

```

### Performance

```{r}
library(pROC)

pred_prob <- predict(glm_presence, type = "response")
obs <- log_reg_data$presence

roc_glm <- roc(obs, pred_prob)
auc(roc_glm)   # 0.7 vs. 0.64 of the hurdle model; slightly better performing 
```

AUC of 0.70 means The habitat variables explain some presence–absence structure, but there is still a lot of overlap/noise in the data.

```{r include=FALSE}
# confusion matrix 
pred_class <- ifelse(pred_prob > 0.5, 1, 0)
table(Predicted = pred_class, Observed = obs)
```

**Confusion Matrix**

| Term                     | Meaning                              | Your value |
| ------------------------ | ------------------------------------ | ---------: |
| True Negatives (TN)  | correctly predicted absence          |       7991 |
| False Negatives (FN)| missed real presences                |        585 |
| False Positives (FP) | predicted presence but none observed |          7 |
| True Positives (TP)  | correctly predicted presence         |          5 |

### Add random effect for HFC/LFC 

```{r}
library(glmmTMB)

glmm_presence <- glmmTMB(
  presence ~ small_woody + depth + velocity + large_woody +
    aquatic_veg + overhanging_veg + cobble_substrate +
    boulder_substrate + undercut_bank +
    (1 | channel_location/location),                     
  data = log_reg_data,
  family = binomial(link = "logit")
)

summary(glmm_presence)


```

**Odds ratio**

```{r}
or_table <- broom.mixed::tidy(glmm_presence, effects = "fixed", conf.int = TRUE) |>
  mutate(
    odds_ratio = exp(estimate),
    or_low = exp(conf.low),
    or_high = exp(conf.high)
  ) |>
  transmute(
    term,
    estimate = round(estimate, 3),
    `Std.Error` = round(std.error, 3),
    `Odds ratio` = round(odds_ratio, 2),
    `95% CI` = sprintf("[%.2f, %.2f]", or_low, or_high),
    p = signif(p.value, 3)
  )

knitr::kable(or_table)

```
```{r}
library(pROC)

pred_prob_glmm <- predict(glmm_presence, type = "response")
obs <- log_reg_data$presence

roc_glmm <- roc(obs, pred_prob_glmm)
auc(roc_glmm) # Area under the curve: 0.8337- much better (for 1|channel_location/location)
#  # Area under the curve: 0.7597 for 1|channel_location

ranef(glmm_presence)$cond$location |>
  as.data.frame()

ranef(glmm_presence)$cond$channel_location |>
  as.data.frame()
# Large variation → occupancy differs strongly by site beyond measured habitat.

```

**Plots**

```{r}
re_loc <- broom.mixed::tidy(glmm_presence, effects = "ran_vals", conf.int = TRUE) |>
  filter(group == "location:channel_location") |>         
  rename(unit = level, re = estimate) |>
  mutate(
    odds_ratio = exp(re),
    or_low  = exp(conf.low),
    or_high = exp(conf.high)
  ) |> 
  arrange(odds_ratio) |>
  mutate(unit = factor(unit, levels = unit))

ggplot(re_loc, aes(x = odds_ratio, y = unit)) +
  geom_vline(xintercept = 1, linetype = 2, color = "gray50") +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = or_low, xmax = or_high), height = 0.2) +
  labs(
    x = "Random-effect odds ratio (relative to global mean)",
    y = NULL,
    title = "Site / channel-unit effects on probability of presence"
  ) +
  scale_x_log10() +
  theme_minimal(base_size = 13)
  


```


### Logistic Regression with Random Effect Discussion 

```{r}

effects_df <- broom::tidy(glm_presence, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(
    odds_ratio = exp(estimate),
    or_low = exp(conf.low),
    or_high = exp(conf.high),
    term = recode(term,
      small_woody = "Small woody debris",
      large_woody = "Large woody debris",
      overhanging_veg = "Overhanging vegetation",
      aquatic_veg = "Aquatic vegetation",
      cobble_substrate = "Cobble substrate",
      boulder_substrate = "Boulder substrate",
      undercut_bank = "Undercut bank",
      depth = "Depth",
      velocity = "Velocity"
    )
  )

ggplot(effects_df, aes(x = odds_ratio, y = reorder(term, odds_ratio))) +
  geom_vline(xintercept = 1, linetype = 2, color = "gray50") +
  geom_point(size = 2.7) +
  geom_errorbarh(aes(xmin = or_low, xmax = or_high), height = 0.2) +
  scale_x_log10() +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL,
    title = "Effect sizes for predictors of fish presence",
    caption = "points > 1 → higher probability of presence; 
    points < 1 → lower probability;  
    wider bars → greater uncertainty"
  ) +
  theme_minimal(base_size = 13)

```


After controlling for local habitat characteristics, the probability of fish occurrence remained significantly structured by site, with LFC consistently supporting higher occupancy and HFC supporting lower occupancy than expected.

Mixed-effects logistic models indicated strong spatial structuring of occurrence beyond the effects of local habitat covariates. Random-intercept estimates for channel units nested within locations showed that some sites exhibited consistently higher baseline probabilities of fish presence, whereas others were characterized by lower occurrence than expected. For example, hatchery ditch (LFC) and junkyard riffle (HFC) had markedly positive random effects (≈2–10× higher odds of presence than the global mean), while eye side channel (LFC) and weir riffle (LFC) showed substantially reduced occurrence (≈60–70% lower odds). These patterns suggest the influence of additional reach-scale or contextual factors not captured by measured microhabitat variables, such as connectivity, geomorphic setting, or broader hydraulic and thermal conditions.

After controlling for other habitat attributes, fish were most likely to occur in units with riparian cover, undercut banks, woody structure, and boulder features, all of which provide refuge and hydraulic complexity. Depth and velocity played secondary roles, suggesting that microhabitat structure is a stronger driver of occurrence than hydraulic conditions alone. Some habitat types (e.g., cobble substrate, aquatic vegetation) did not show clear relationships with presence once other covariates were included.


```{r}
knitr::knit_exit()
```

