---
title: "Feather River - Mini Snorkel Data Modeling and Analysis for Steelhead"
author: "Maddee Wiggins (FlowWest)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3      
    toc_float: true   
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(sjPlot)
library(pROC)
library(glmmTMB)
library(sf)
library(leaflet)
library(scales)

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)

theme_set(
  theme_minimal() +
    theme(
      plot.title = element_text(size = 11, face = "bold"),
      axis.title.x = element_text(size = 14),
      axis.title.y = element_text(size = 14)
    )
)


custom_colors <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

colors_full <-  c("#9A8822", "#F5CDB4", "#F8AFA8", "#FDDDA0", "#74A089", #Royal 2
                  "#899DA4", "#C93312", "#DC863B", # royal 1 (- 3)
                  "#F1BB7B", "#FD6467", "#5B1A18", "#D67236",# Grand Budapest 1 (-4)
                  "#D8B70A", "#02401B", "#A2A475", # Cavalcanti 1
                  "#E6A0C4", "#C6CDF7", "#D8A499", "#7294D4", #Grand Budapest 2
                  "#9986A5", "#EAD3BF", "#AA9486", "#B6854D", "#798E87", # Isle of dogs 2 altered slightly
                  "#F3DF6C", "#CEAB07", "#D5D5D3", "#24281A", # Moonriese 1, 
                  "#798E87", "#C27D38", "#CCC591", "#29211F", # moonrise 2
                  "#85D4E3", "#F4B5BD", "#9C964A", "#CDC08C", "#FAD77B" ) # moonrise 3 

options(scipen=999)

```

## Objective

Develop a model that reflects the significance of cover, substrate,
depth, and velocity on steelhead presence and absence.

## Analysis Overview

Hurdle model explored but was a very poor fit for the count. Excluding
from report.

## Review Data

```{r}
source(here::here("data-raw", "pull_from_edi.R"))

mini_snorkel_raw <- mini_fish_raw |> 
  left_join(mini_locations_raw |> distinct()) |> 
  mutate(count = ifelse(is.na(count), 0, count),
         fish_presence = as.factor(ifelse(count < 1, "0", "1")),
         month = month(date)) |> 
  filter(species %in% c("steelhead trout (wild)", "steelhead trout (clipped)") | count == 0) |> 
  glimpse()

```

## Data Exploration

### High Flow vs. Low Flow Channel

Most of the steelhead were found in the Low Flow channel.

```{r echo=FALSE, message=FALSE, warning=FALSE}
mini_snorkel_raw |> 
  filter(count > 0) |> 
  ggplot() + 
  geom_histogram(aes(count, fill = channel_location), binwidth = 50) + 
  scale_fill_manual(values = custom_colors) + 
  ylab("steelhead count") +
  ggtitle('Figure 1. Histogram of steelhead count in the high and low flow channels') +facet_grid(~channel_location)

mini_snorkel_model_ready <- mini_snorkel_raw
```

```{r}

# looking at overall count data between the low flow/high flow channels
mini_snorkel_model_ready |> 
  group_by(channel_location) |> 
  tally(count) |> 
  knitr::kable(caption = "Table 1. Total count of chinook salmon between high flow and low flow channels")

mini_snorkel_model_ready |> 
  group_by(channel_location) |> 
  summarize(n_sites = length(unique(location))) |> 
  knitr::kable(caption = "Table 2. Number of sampling sites in the high flow and low flow channels")
```

### Temporal exploration

The majority of fish are found in the low flow channel with the highest
counts in April.

```{r echo=FALSE, message=FALSE, warning=FALSE}

mini_snorkel_model_ready |> 
  group_by(channel_location, month = lubridate::month(date)) |> 
  tally(count) |> 
  ggplot() + 
  geom_col(aes(x = as.factor(month), y = n, fill = channel_location), position = "dodge") + 
  ylab("fish count") +
  xlab("month") +
  ggtitle('Figure 2. Steelhead presence as a function of month and high flow/low flow channel') + 
  scale_fill_manual(values = custom_colors)

```

### Redd Data Integration and Exploration

**Data Question:** - for sites that include roving and transect, can I
aggregate or is that double counting? Currently grouped by date and
location name to get a total number of redds.

```{r}
redd_data_raw <- readxl::read_excel(here::here("data-raw", "SH Redd Survey.xlsx")) |> 
  janitor::clean_names() 

# remove zero redds
redd_data_with_redds <- redd_data_raw |> 
  rename(number_redds = number_of_redds) |> 
  filter(number_redds > 0) |> 
  mutate(lat = as.numeric(lat),
         long = as.numeric(long)) |>  
  filter(!is.na(lat) | !is.na(long))
```

**Redd data over time**

This visual represents the number of redds over time at each location.
It helps provide context on which sites generally have redds and if they
have redds consistently over time.

```{r}
redd_data_with_redds |> 
  group_by(date, location) |> 
  summarise(total_redds = sum(number_redds)) |> 
  ggplot(aes(x = lubridate::year(date), y = location, fill = total_redds)) +
  geom_tile(color = "white",  linewidth = 0.5) + 
  xlab("Year")
```

Total number of redds by year

```{r}
redd_by_year <- redd_data_with_redds |> group_by(year = year(date)) |> summarise(total_redds = sum(number_redds)) 

knitr::kable(redd_by_year)

```

Combine redd data with mini snorkel to see if redds are a spatial
indicator or spawning potential/habitat quality.

```{r}
redd_data_by_location <- redd_data_with_redds |> 
  rename(date_redd = date) |> 
  filter(!is.na(lat)) |> 
  sf::st_as_sf(coords = c("long", "lat"), crs = 4326)

# redd_data_by_location |> 
#   ggplot(aes(x = location, y = n_redds)) +
#   geom_col() +
#   coord_flip()

mini_sf <- mini_snorkel_model_ready |> 
  filter(!is.na(longitude)) |> 
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) 
```

Find the nearest transect location from the Mini Snorkel data to each
redd

```{r}
dist_to_nearest <- st_distance(
  redd_data_by_location,
  mini_sf[st_nearest_feature(redd_data_by_location, mini_sf), ],
  by_element = TRUE
)

redd_near <- redd_data_by_location |>
  mutate(dist_to_nearest = as.numeric(dist_to_nearest)) |> 
  filter(dist_to_nearest <= 50)

redd_near_with_snorkel <- st_join(
  redd_near |> select(-location),
  mini_sf,
  join = st_nearest_feature
)
```

Redds are joined to the nearest mini snorkel transect location within
50-meters. The following histogram shows the distribution of the redd
distances; 50-meters was chosen based off of the high counts (shown as
the red dashed line). Note that three outliers were removed, shown in
the table below. The locations (which are visible as light blue circles
in the leaflet map) look to be errors.

```{r}
redd_outliers <- redd_data_by_location |>
  mutate(dist_to_nearest = as.numeric(dist_to_nearest)) |> 
  filter(dist_to_nearest > 1000)

knitr::kable(redd_outliers |> st_drop_geometry() |> select(date = date_redd, location, number_redds), caption = "Outliers - The following three data points seem to have incorrect coordinates")

redd_data_by_location |>
  mutate(dist_to_nearest = as.numeric(dist_to_nearest)) |> 
  filter(dist_to_nearest < 1000) |> # removing outliers
  ggplot(aes(x = dist_to_nearest)) +
  geom_histogram(binwidth = 5, boundary = 0) +
  geom_vline(xintercept = 50, linetype = 2, color = "red") +
  labs(
    title = "Distribution of nearest-neighbor distances"
  ) +
  theme_minimal(base_size = 13)

redd_summary <- redd_near_with_snorkel |> 
  st_drop_geometry() |> 
  group_by(location) |> 
  summarise(
    redd_total = sum(number_redds),
    redd_mean = mean(number_redds),
    redd_median = median(number_redds),
    redd_presence = as.integer(redd_total > 0)
  )
```

A visual representation of the amount of redds joined with each of the
mini snorkel transect locations. Outliers that have been removed are
shown in blue.

```{r}

locations_sf <- sf::st_as_sf(mini_locations_raw |> 
                               distinct(location_table_id, date, channel_location, .keep_all = T) |> 
                               na.omit(), 
                             coords = c('longitude', 'latitude')) |> 
  select(location_table_id, location, channel_type, date) |> 
  mutate(location_table_id = paste0(date, " (location id: ", location_table_id, ")")) |> 
  mutate(color = colors_full[match(location, location)]) |> 
  group_by(location) |> 
  slice(1) |> # This is a bit of a workaround since there are multiple lat/longs for reach location that are near eachother. Since this is just for a visual, I think it's fine. 
  ungroup()

redd_summary_sf <- redd_summary |> 
  left_join(locations_sf) |> 
  select(location, redd_presence, redd_total, geometry) |> 
  distinct() |> 
  sf::st_as_sf(sf_column_name = "geometry", crs = 4326)

leaflet() |>
  addProviderTiles(providers$Esri.WorldImagery, group = "Aerial Imagery") |>
  addProviderTiles(providers$OpenStreetMap, group = "Street Map") |>
  addCircleMarkers(data = redd_outliers,
                   popup = paste0("Outlier Location Name: ", redd_outliers$location,"<br>",
                                  "Date: ", redd_outliers$date_redd)) |> 
   addCircleMarkers(data = locations_sf, 
                     layerId = ~location, 
                     color = ~color,
                     radius = 6,
                     fill = TRUE,
                     fillOpacity = 0.2,
                     opacity = 0.6,
                     popup = paste0("Location Name: ", locations_sf$location,
                                    "<br>",
                                    "Channel Type: ", locations_sf$channel_type)) |> 
  addCircleMarkers(data = redd_summary_sf, 
                   color = "darkred",
                   radius = ~rescale(redd_total, to = c(3, 20)),
                   fill = TRUE,
                   fillOpacity = 1,
                   opacity = 1, 
                   popup = paste0("Location Name: ", redd_summary_sf$location,
                                    "<br>",
                                    "N Redds: ", redd_summary_sf$redd_total)) |> 
   
  addLayersControl(
    baseGroups = c("Aerial Imagery", "Street Map"),
    position = "topleft",
    options = layersControlOptions(collapsed = FALSE)
  ) |> 
  addLegend(colors = "darkred", labels = "Redd Locations") 

```

### Variables of Interest

The variables of interest include cover, substrate, velocity and depth
variables known to be important for steelhead rearing habitat. We are
also including the number of redds found near each mini snorkel transect
and whether or not there were redds nearby.

-   Velocity - numeric
-   Depth - numeric
-   Number of Redds - numeric
-   Redd at location - 0/1
-   Month - categorical (3-8)

Substrate and cover variables are measured as percentages are converted
to binary presence/absence (1/0) by establishing a threshold percentage
of 20%. Overhanging vegetation was measured at 1/2 meter overhead and
more than 1/2 meter overhead. These categories were combined for
simplicity and for comparison with other studies, such as those by Mark
Gard.

-   Undercut bank - 0/1
-   Aquatic vegetation - 0/1
-   Overhanging vegetation - 0/1
-   Small woody cover - 0/1
-   Large woody cover - 0/1
-   Boulder substrate - 0/1
-   Cobble substrate - 0/1

## Build Model Data

All cover variables were converted to presence/absence using a threshold
of 20%. The following is the data structure of the model input data.

```{r}

percent_threshold <- 20

# use a hurdle model (Gard 2024); 
# pre-processing - all numeric values must be rounded to whole numbers
model_data <- mini_snorkel_model_ready |>
  select(count, location, channel_location, depth, velocity, contains("inchannel"), contains("overhead"), "percent_cobble_substrate", "percent_boulder_substrate", "percent_undercut_bank", month, fl_mm, channel_geomorphic_unit, reach_length, reach_width, channel_type, month) |>
  # create new cover variables, based off Gard 2023
  mutate(small_woody = percent_small_woody_cover_inchannel,
         large_woody = percent_large_woody_cover_inchannel,
         boulder_substrate = percent_boulder_substrate,
         cobble_substrate = percent_cobble_substrate, 
         undercut_bank = percent_undercut_bank, 
         aquatic_veg = percent_submerged_aquatic_veg_inchannel, 
         overhanging_veg = percent_cover_half_meter_overhead + percent_cover_more_than_half_meter_overhead) |>
  # transform to presence/absence based on a defined threshold
  mutate(cobble_substrate = ifelse(percent_cobble_substrate >= percent_threshold, 1, 0 ),
         boulder_substrate = ifelse(percent_boulder_substrate >= percent_threshold, 1, 0 ),
         small_woody = ifelse(small_woody >= percent_threshold, 1, 0 ),
         large_woody = ifelse(large_woody >= percent_threshold, 1, 0 ),
         aquatic_veg = ifelse(aquatic_veg >= percent_threshold, 1, 0),
         undercut_bank = ifelse(undercut_bank >= percent_threshold, 1, 0),
         no_cover_overhead = ifelse(percent_no_cover_overhead >= percent_threshold, 1, 0),
         overhanging_veg = ifelse(overhanging_veg >= percent_threshold, 1, 0)) |>
  #remove no cover variables (no cover overhead, no cover in channel)
  select(-contains("no_cover")) |> 
  distinct() |> 
  mutate(fl_mm = ifelse(is.na(fl_mm), 0, fl_mm)) |> 
  na.omit() |> 
  select(-fl_mm) |> 
  left_join(redd_summary) |> 
  mutate(
    redd_total = replace_na(redd_total, 0),
    redd_presence = replace_na(redd_presence, 0),
    month = as.factor(month), 
  ) |> 
  glimpse()

```

## Model Performance Evaluation Overview

Model performance was evaluated using a combination of receiver
operating characteristic (ROC) analysis and confusion matrix–based
classification metrics. The area under the ROC curve (AUC) was used to
assess the model’s ability to discriminate between presence and absence
across all possible probability thresholds. AUC provides a
threshold-independent measure of performance, where values near 0.5
indicate no discrimination and higher values indicate increasing ability
to correctly rank presences above absences.

In addition, model predictions were converted to binary classifications
using a fixed probability threshold, and a confusion matrix was used to
summarize agreement between predicted and observed outcomes. The
confusion matrix tabulates true positives, true negatives, false
positives, and false negatives, allowing evaluation of classification
behavior such as sensitivity to presences and specificity to absences.
Because the dataset was imbalanced, with many more absences than
presences, emphasis was placed on interpreting the confusion matrix in
conjunction with AUC rather than relying on overall accuracy alone.

Confusion matrises have the following layout where "True" fish presences
are represented as `val1` and `val4`. "False" positives are represented
by `val2` as cases where fish presence was predicted but not observed.
"False" negatives are represented by `val3`.

|                        | Observed Absence (0) | Observed Presence (1) |
|------------------------|----------------------|-----------------------|
| Predicted Absence (0)  | `val1`               | `val3`                |
| Predicted Presence (1) | `val2`               | `val4`                |

Together, these metrics provide complementary perspectives on model
performance: AUC characterizes overall discriminatory ability
independent of threshold choice, while the confusion matrix illustrates
how predictions behave at a specific cutoff and highlights trade-offs
between detecting presences and avoiding false positives.

## Logistic Regression Approach

Because fish counts were highly zero-inflated and exhibited extreme
variability among non-zero values, abundance models performed poorly. We
therefore focused on modeling fish occurrence using logistic regression,
which better matches the information content of the data and provides
more reliable inference on habitat associations.

```{r}
log_reg_data <- model_data |> 
  mutate(presence = as.integer(count > 0))
```

### Build Logistic Regression Model

A simple logistic regression to start.

```{r}
glm_presence <- glm(
  presence ~ small_woody + depth + velocity + large_woody +
    aquatic_veg + overhanging_veg + cobble_substrate +
    boulder_substrate + undercut_bank + redd_total + redd_presence,
  data = log_reg_data,
  family = binomial(link = "logit")
)

summary(glm_presence)

```

### Performance

```{r}

pred_prob <- predict(glm_presence, type = "response")
obs <- log_reg_data$presence

roc_glm <- roc(obs, pred_prob)
auc(roc_glm)   # slightly better performing 

# pred_class <- ifelse(pred_presence > 0.5, 1, 0)
# table(Predicted = pred_class, Observed = obs_presence)
```

AUC of `r auc(roc_glm)` means The habitat variables explain some
presence–absence structure, but there is still a lot of overlap/noise in
the data.

Confusion Matrix:

```{r echo=FALSE}
# confusion matrix 
pred_class <- ifelse(pred_prob > 0.5, 1, 0)
table(Predicted = pred_class, Observed = obs)
```

### Logistic Regression with Random Effect

### Random Effect of Location

Using the high flow, low flow channel as the random effect:

```{r}

glmm_presence_location <- glmmTMB(
  presence ~ small_woody + depth + velocity + large_woody +
    aquatic_veg + overhanging_veg + cobble_substrate +
    boulder_substrate + undercut_bank + redd_total + redd_presence + 
    (1 | location),                     
  data = log_reg_data,
  family = binomial(link = "logit")
)

summary(glmm_presence_location)

pred_prob_glmm_loc <- predict(glmm_presence_location, type = "response")
obs <- log_reg_data$presence

roc_glmm_loc <- roc(obs, pred_prob_glmm_loc)
auc(roc_glmm_loc) 

# confusion matrix 
pred_class_loc <- ifelse(pred_prob_glmm_loc > 0.5, 1, 0)
table(Predicted = pred_class_loc, Observed = obs)

```

### Random Effect of Channel Location \| Location

Nesting the transect location within the high flow/low flow channel as
the random effect:

```{r}

glmm_presence <- glmmTMB(
  presence ~ small_woody + depth + velocity + large_woody +
    aquatic_veg + overhanging_veg + cobble_substrate +
    boulder_substrate + undercut_bank + redd_total + redd_presence + 
    (1 | channel_location/location),                     
  data = log_reg_data,
  family = binomial(link = "logit")
)

summary(glmm_presence)

```

```{r}
pred_prob_glmm <- predict(glmm_presence, type = "response")
obs <- log_reg_data$presence

roc_glmm <- roc(obs, pred_prob_glmm)
auc(roc_glmm) 

```

```{r echo=FALSE}
# confusion matrix 
pred_class <- ifelse(pred_prob_glmm > 0.5, 1, 0)
table(Predicted = pred_class, Observed = obs)
```

### Random Effect of Location and Month

Using location and month as the random effect.

Based off of the data exploration with majority of steelhead in the low
flow channel, it is unsurprising that location is a better predictor.

```{r}

glmm_presence_location_month <- glmmTMB(
  presence ~ small_woody + depth + velocity + large_woody +
    aquatic_veg + overhanging_veg + cobble_substrate +
    boulder_substrate + undercut_bank + redd_total + redd_presence + 
    (1 | location) + (1 | month),                     
  data = log_reg_data,
  family = binomial(link = "logit")
)

summary(glmm_presence_location_month)

pred_prob_glmm_loc_month <- predict(glmm_presence_location_month, type = "response")
obs <- log_reg_data$presence

roc_glmm_loc_month <- roc(obs, pred_prob_glmm_loc_month)
auc(roc_glmm_loc_month) 

# confusion matrix 
pred_class_loc_month <- ifelse(pred_prob_glmm_loc_month > 0.5, 1, 0)
table(Predicted = pred_class_loc_month, Observed = obs)

```

### Figures

The following figure represents the overall effect of month on fish
presence while keeping all other months constant. Points greater than 1
represent a higher probability of fish presence and points less than 1
represent a lower probability. Wider bars represent greater uncertainty.

```{r}

month_effects_df <-  broom.mixed::tidy(glmm_presence_location_month, effects = "ran_vals", conf.int = TRUE) |>
  filter( group == "month") |> 
  rename(unit = level, re = estimate) |>
  mutate(
    odds_ratio = exp(re),
    or_low  = exp(conf.low),
    or_high = exp(conf.high)
  ) |>
  arrange(desc(unit)) |>
  distinct() |>
  mutate(unit = factor(unit, levels = unit)) 

ggplot(month_effects_df, aes(x = odds_ratio, y = unit)) +
  geom_vline(xintercept = 1, linetype = 2, color = "gray50") +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = or_low, xmax = or_high), height = 0.2) +
  labs(
    x = "Random-effect odds ratio (relative to global mean)",
    y = NULL,
    title = "Month effects on probability of presence of steelhead"
  ) +
  scale_x_log10() +
  theme_minimal(base_size = 13)

```

The following figure represents the overall effect of each of the sites
on fish presence while keeping all other sites constant. Points greater
than 1 represent a higher probability of fish presence and points less
than 1 represent a lower probability. Wider bars represent greater
uncertainty.

Larger impact for low flow channel.

```{r}
re_loc <- broom.mixed::tidy(glmm_presence_location_month, effects = "ran_vals", conf.int = TRUE) |>
  filter(group == "location") |>
  rename(unit = level, re = estimate) |>
  left_join(mini_snorkel_raw |> select(unit = location, channel_location),
            by = "unit") |>
  mutate(
    odds_ratio = exp(re),
    or_low  = exp(conf.low),
    or_high = exp(conf.high)
  ) |>
  arrange(odds_ratio) |>
  distinct() |>
  mutate(unit = factor(unit, levels = unit)) 

ggplot(re_loc, aes(x = odds_ratio, y = unit, color = channel_location)) +
  geom_vline(xintercept = 1, linetype = 2, color = "gray50") +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = or_low, xmax = or_high), height = 0.2) +
  labs(
    x = "Random-effect odds ratio (relative to global mean)",
    y = NULL,
    title = "Site effects on probability of presence"
  ) +
  scale_color_manual(
    values = c(
      "HFC" = "#1f78b4",   # blue
      "LFC" = "#33a02c"    # green
    ),
    labels = c(
      "HFC" = "High-flow channel",
      "LFC" = "Low-flow channel"
    ),
    name = "Channel type"
  ) +
  scale_x_log10() +
  theme_minimal(base_size = 13)




```

The following figure represents the overall effect of each of the
predictors for fish presence when keeping all other predictors constant.
Points greater than 1 represent a higher probability of fish presence
and points less than 1 represent a lower probability. Wider bars
represent greater uncertainty. The magnitude of these effects is modest
on a per-unit basis, but their significance reflects the precision of
the estimates and the large sample size rather than a lack of biological
relevance.

```{r}

effects_df <- broom::tidy(glmm_presence_location_month, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  filter(term != "sd__(Intercept)") |> 
  mutate(
    odds_ratio = exp(estimate),
    or_low = exp(conf.low),
    or_high = exp(conf.high),
    term = recode(term,
                  small_woody = "Small woody debris*",
                  large_woody = "Large woody debris",
                  overhanging_veg = "Overhanging vegetation*",
                  aquatic_veg = "Aquatic vegetation*",
                  cobble_substrate = "Cobble substrate",
                  boulder_substrate = "Boulder substrate",
                  undercut_bank = "Undercut bank",
                  depth = "Depth",
                  velocity = "Velocity*",
                  redd_total = "Total Number or Redd*",
                  redd_presence = "Redds Nearby*"
    )
  )

ggplot(effects_df, aes(x = odds_ratio, y = reorder(term, odds_ratio))) +
  geom_vline(xintercept = 1, linetype = 2, color = "gray50") +
  geom_point(size = 2.7) +
  geom_errorbarh(aes(xmin = or_low, xmax = or_high), height = 0.2) +
  scale_x_log10() +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL,
    title = "Effect sizes for predictors of steelhead presence", 
    caption = "* represents model significance p<0.05"
  ) +
  theme_minimal(base_size = 13)

```

### Discussion

```{r}
knitr::knit_exit()
```
